---
title: "Classification examples"
output: html_notebook
---

# Background


# Setup libs and environment

```{r}
library(tidyverse)
library(keras)
library(brms)
library(janitor)
library(kableExtra)
library(corrplot)
library(caret)
library(pscl)
ggplot2::theme_set(theme_classic(base_size = 18))
options(scipen = 999999)
knitr::opts_chunk$set(fig.width = 5, fig.height = 3, dpi = 300)
```

# Titanic dataset

Taken from https://www.kaggle.com/code/dmilla/introduction-to-decision-trees-titanic-dataset

## Load the data

```{r}
titanic <- read_csv("input/titanic/titanic.csv") |>
  clean_names()
```

## Data exploration

### Show me what you got

Each row shows a single passenger, if they survived, their class, name, sex, age, no. siblings, N. parents/childred, ticket number & fare cost.

```{r}

titanic |>
  head() |>
  kbl() |>
  kable_classic()

```

## Data structure

* Lots of missing age data.
* Almost no cabin data.
* One missing fare and one missing embarked.

```{r}
return_dtype <- function(x) {
  if(is.na(x)) {
    return("NA")
  }
  return(typeof(x))
}
return_dtype_vector <- function(x) {
  sapply(x, return_dtype)
}

titanic |>
  mutate(across(.fns = return_dtype_vector)) |>
  mutate(row_num = row_number()) |>
  pivot_longer(cols = !row_num) |>
  mutate(name = factor(name,
                       levels = names(titanic),
                       )) |>
  ggplot(aes(x = name, y = row_num, fill = value)) +
  geom_tile() +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## How many survived

* More survived than died

```{r}
titanic |>
  group_by(survived) |>
  summarise(n = n()) |>
  ggplot(aes(x = as.factor(survived), y = n)) +
  geom_bar(stat = "identity")

titanic |>
  group_by(survived) |>
  summarise(n = n()) |>
  ungroup() |>
  mutate(t = sum(n),
         survived_percent = round(n/t*100, 2)
         ) |>
  select(survived, survived_percent) |>
  kbl() |>
  kable_classic()
```

## Gender distribution

### Total

* More males onboard

```{r}
titanic |>
  group_by(sex) |>
  summarise(n = n()) |>
  ggplot(aes(x = sex, y = n)) +
  geom_bar(stat = "identity")
```

### Survived

* More males died
* Higher & of males died

```{r}
titanic |>
  group_by(sex, survived) |>
  summarise(n = n()) |>
  ggplot(aes(x = sex, y = n, fill = as.factor(survived))) +
  geom_bar(stat = "identity", position = position_dodge(1))

titanic |>
  group_by(sex, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(sex) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  ggplot(aes(x = sex, y = percent_died, fill = as.factor(survived))) +
  geom_bar(stat = "identity")

```

## Class distribution

### Total

* Equal class 1 & 2, more class 3

```{r}
titanic |>
  group_by(pclass) |>
  summarise(n = n()) |>
  ggplot(aes(x = pclass, y = n)) +
  geom_bar(stat = "identity")
```


### Survived

* As expected, higher class passengers have a higher % chance of survival

```{r}
titanic |>
  group_by(pclass, survived) |>
  summarise(n = n()) |>
  ggplot(aes(x = pclass, y = n, fill = as.factor(survived))) +
  geom_bar(stat = "identity", position = position_dodge(1))

titanic |>
  group_by(pclass, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(pclass) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  ggplot(aes(x = pclass, y = percent_died, fill = as.factor(survived))) +
  geom_bar(stat = "identity")

```

## Numbers at age

### Total

* 263 missing ages, sad

```{r}
titanic |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 1)
```

### Survived

* Really interesting non-linear interaction with age here.
* Probably too many to impute. Will have to decide based on model accuracy if we should include this or not.

```{r}
titanic |>
  mutate(age = round(age, 0)) |>
  group_by(age, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(age) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  filter(survived == 0) |>
  ggplot(aes(x = age, y = percent_died)) +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", colour = "red") +
  geom_line()
```


## Number siblings/spouces

### Total

* Not a lot in higher sibling counts

```{r}
titanic |>
  ggplot(aes(x = sib_sp)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(0, 9, 1), labels = seq(0, 9, 1))
```


### Survived

* Looks very good, but probably due to low samples in higher sibling counts
* Limiting it to 2 siblings shows strong fit, but this could be correlated with other things. Lets see later on.

```{r}
titanic |>
  # mutate(age = round(age, 0)) |>
  group_by(sib_sp, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(sib_sp) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  filter(survived == 0) |>
  ggplot(aes(x = sib_sp, y = percent_died)) +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", colour = "red") +
  geom_line()


titanic |>
  mutate(sib_sp = ifelse(sib_sp>=2, 2, sib_sp)) |>
  group_by(sib_sp, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(sib_sp) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  filter(survived == 0) |>
  ggplot(aes(x = sib_sp, y = percent_died)) +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", colour = "red") +
  geom_line()
```


## Number parents/children

### Total

```{r}
titanic |>
  ggplot(aes(x = parch)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(0, 9, 1), labels = seq(0, 9, 1))
```

### Survived

* Same as siblings/spoices onboard, low sample sizes show good fit
* Lets check for correlation with other things before incusion
* Could be a good fit though

```{r}
titanic |>
  # mutate(age = round(age, 0)) |>
  group_by(parch, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(parch) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  filter(survived == 0) |>
  ggplot(aes(x = parch, y = percent_died)) +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", colour = "red") +
  geom_line()

titanic |>
  mutate(parch = ifelse(parch>=2, 2, parch)) |>
  group_by(parch, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(parch) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  filter(survived == 0) |>
  ggplot(aes(x = parch, y = percent_died)) +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", colour = "red") +
  geom_line()
```


## Fare distribution

### Total

```{r}
titanic |>
  ggplot(aes(x = fare)) +
  geom_histogram(binwidth = 25)
```

### Survived

* Fare, as expected, shows a good relationship with % died
* Capping at £100 will help with low sample sizes for larger fares
* Will correlate with class, easier to use these continuous variables comapred to factors in models

```{r}
titanic |>
  mutate(fare = round(fare, 0)) |>
  group_by(fare, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(fare) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  filter(survived == 0) |>
  ggplot(aes(x = fare, y = percent_died)) +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", colour = "red") +
  geom_line()

titanic |>
  mutate(fare = round(fare, 0)) |>
  mutate(fare = ifelse(fare>=100, 100, fare)) |>
  group_by(fare, survived) |>
  summarise(n = n()) |>
  ungroup() |>
  group_by(fare) |>
  mutate(t = sum(n)) |>
  ungroup() |>
  mutate(percent_died = n/t*100) |>
  filter(survived == 0) |>
  ggplot(aes(x = fare, y = percent_died)) +
  geom_smooth(se = F) +
  geom_smooth(se = F, method = "lm", colour = "red") +
  geom_line()
```

## Correlation plots

```{r}


cor_matrix <- titanic |>
  filter(!is.na(fare)) |>
  filter(!is.na(age)) |>
  mutate(sex = ifelse(sex == "male", 1, 0)) |>
  select(survived, pclass, sex, age, sib_sp, parch, fare) |>
  cor()

cor_matrix |>
  round(2) |>
  head() |>
  kbl() |>
  kable_classic()

cor.mtest <- function(mat, ...) {
    mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat<- matrix(NA, n, n)
    diag(p.mat) <- 0
    for (i in 1:(n - 1)) {
        for (j in (i + 1):n) {
            tmp <- cor.test(mat[, i], mat[, j], ...)
            p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
        }
    }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}
# matrix of the p-value of the correlation
p.mat <- titanic |>
  filter(!is.na(fare)) |>
  filter(!is.na(age)) |>
  mutate(sex = ifelse(sex == "male", 1, 0)) |>
  select(survived, pclass, sex, age, sib_sp, parch, fare) |>
  cor.mtest()

corrplot(cor_matrix, type="upper", order="hclust", 
         p.mat = p.mat, sig.level = 0.01)
```


## Data exploration lessons learned

* Inclusion of sex, fare & parch essential
* Age has interesting non-liner, but lots of missing data so may be worth not including it
* sib_sp may have some non-linear relationship

# Build models

## Test train split

```{r}
test_id <- createDataPartition(titanic$survived,
                               p = 0.8,
                               list = F)
test <- titanic[-test_id,]
train <- titanic[test_id,]
```


## Frequentist logistic regression

### Build models

```{r}
freq_m1 <- glm(
  survived ~
    1,
  data = train,
  family = binomial()
)

summary(freq_m1)

freq_m2 <- glm(
  survived ~
    sex,
  data = train,
  family = binomial()
)

summary(freq_m2)

freq_m3 <- glm(
  survived ~
    sex +
    fare,
  data = train,
  family = binomial()
)

summary(freq_m3)

freq_m4 <- glm(
  survived ~
    sex +
    fare +
    parch,
  data = train,
  family = binomial()
)

summary(freq_m4)

freq_m5 <- glm(
  survived ~
    sex +
    fare +
    parch +
    age,
  data = train,
  family = binomial()
)

summary(freq_m5)

freq_m6 <- glm(
  survived ~
    sex +
    fare +
    parch +
    age +
    sib_sp,
  data = train,
  family = binomial()
)

```

### Accuracy

```{r}

test_accuracy <- function(model, test_data) {
  
  res <- predict(model, newdata = test_data, type = "response") |>
    unname()
  
  test_data <- test_data |>
    mutate(fitted = ifelse(res > 0.5, 1, 0)) |>
    filter(!is.na(fitted))
  
  error <- mean(test_data$fitted != test_data$survived)
  
  return(1-error)
  
}

```


