---
title: "Classification examples"
output: html_notebook
---

# Background


# Setup libs and environment

```{r}
library(tidyverse)
library(keras)
library(brms)
library(janitor)
library(kableExtra)
ggplot2::theme_set(theme_classic(base_size = 18))
options(scipen = 999999)
```

# Titanic dataset

Taken from https://www.kaggle.com/code/dmilla/introduction-to-decision-trees-titanic-dataset

## Load the data

```{r}
titanic <- read_csv("input/titanic/titanic.csv") |>
  clean_names()
```

## Data exploration

### Show me what you got

Each row shows a single passenger, if they survived, their class, name, sex, age, no. siblings, N. parents/childred, ticket number & fare cost.

```{r}

titanic |>
  head() |>
  kbl() |>
  kable_classic()

```

## Data structure

* Lots of missing age data.
* Almost no cabin data.
* One missing fare and one missing embarked.

```{r, fig.width=5, fig.height=3, dpi = 300}
return_dtype <- function(x) {
  if(is.na(x)) {
    return("NA")
  }
  return(typeof(x))
}
return_dtype_vector <- function(x) {
  sapply(x, return_dtype)
}

titanic |>
  mutate(across(.fns = return_dtype_vector)) |>
  mutate(row_num = row_number()) |>
  pivot_longer(cols = !row_num) |>
  mutate(name = factor(name,
                       levels = names(titanic),
                       )) |>
  ggplot(aes(x = name, y = row_num, fill = value)) +
  geom_tile() +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## How many survived

* More survived than died

```{r}
titanic |>
  group_by(survived) |>
  summarise(n = n()) |>
  ggplot(aes(x = as.factor(survived), y = n)) +
  geom_bar(stat = "identity")

titanic |>
  group_by(survived) |>
  summarise(n = n()) |>
  ungroup() |>
  mutate(t = sum(n),
         survived_percent = round(n/t*100, 2)
         ) |>
  select(survived, survived_percent) |>
  kbl() |>
  kable_classic()
```

## Gender distribution

### Total

* More males onboard

```{r}
titanic |>
  group_by(sex) |>
  summarise(n = n()) |>
  ggplot(aes(x = sex, y = n)) +
  geom_bar(stat = "identity")
```

### Survived

* More males died
* Higher & of males died

```{r}
titanic |>
  group_by(sex, survived) |>
  summarise(n = n()) |>
  ggplot(aes(x = sex, y = n, fill = as.factor(survived))) +
  geom_bar(stat = "identity", position = position_dodge(1))

titanic |>
  group_by(sex, survived) |>
  summarise(n = n()) |>
  ggplot(aes(x = sex, y = n, fill = as.factor(survived))) +
  geom_bar(stat = "identity", position = position_dodge(1))

```

## Class distribution

### Total
### Survived

## Numbers at age

### Total
### Survived

## Nuber siblings

### Total
### Survived

## Number parents

### Total
### Survived

## Fare distribution

### Total
### Survived

## Embarked

### Total
### Survived
